---
title: "Caso de estudio 1 Ciclystic"
author: "Ronal Castellanos"
date: "2023-07-12"
output:
  pdf_document: default
  html_document: default
editor_options:
  chunk_output_type: console
---

# About the company.
In 2016, Cyclistic launched a successful bike share offering. Since then, the program grew to a fleet of 5,824 geo-tagged and locked bikes at a network of 692 stations across Chicago. Bikes can be unlocked from one station and returned to any other station in the system at any time.

Until now, Cyclistic's marketing strategy was based on building overall brand recognition and appealing to broad consumer segments. One of the approaches that helped make this possible was the flexibility of its pricing plans: single-ride passes, full-day passes, and annual memberships. Customers who purchase single-ride passes or full-day passes are referred to as occasional riders. Customers who purchase annual memberships are called Cyclistic members.

Cyclistic's financial analysts concluded that annual memberships are much more profitable than occasional riders. While flexible pricing helps Cyclistic attract more customers, Moreno believes that maximizing the number of annual members will be key to future growth. Rather than creating a marketing campaign that targets all new customers, Moreno believes there is plenty of opportunity to convert casual cyclists into members. She points out that casual cyclists already know about Cyclistic's program and have chosen Cyclistic for their mobility needs.

Moreno set a clear goal: Design marketing strategies aimed at converting casual cyclists into annual members. However, to do that, the marketing analyst team needs to better understand how annual members and casual cyclists differ, why casual cyclists would buy a membership, and how digital media might affect their marketing tactics. Moreno and his team are interested in analyzing Cyclistic's historical bike trip data to identify trends.

Translated with www.DeepL.com/Translator (free version)

## Data Source
This data was obtained from from https://divvy-tripdata.s3.amazonaws.com/index.html
Data obtained for June 2022 to June 2023

### Step 1: Prepare the files
1. First row removed from each .csv file, just one keep first header row
2. all files merged by cmd comand copy *.csv, ciclystic.csv


### Step 2: Load packages
Packages intalled:

install.packages("tidyverse")
install.packages("skirm")
install.packages("skimr")
install.packages("janitor")

#### Step 2.1: Loading packages
```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}

library(tidyverse)
library(skimr)
library(janitor)
library(dplyr)
library(lubridate)
knitr::opts_chunk$set(tidy = TRUE, tidy.opts = list(comment = FALSE))
```


### Step 3: Import and Clean data

Data loadd from .csv file merged.
```{r load dataset, echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
rides<- read_csv("ciclystic.csv")
```

#### 3.1 Filter null values
```{r remove null values, echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}

cleaned_rides <- na.omit(rides)


```

#### 3.2 Convert datetime to same format

There is two types of dateformat in the file YYYY-mm-dd HH:MM:SS y MM/DD/YYY HH:MM
They were matched to datetimestamp format
```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
          
trips <- cleaned_rides %>%
  mutate(started_at = parse_date_time(started_at, orders = c("%Y-%m-%d %H:%M:%S", "%Y-%m-%d %H:%M", "%m/%d/%Y %H:%M")))%>%
   mutate(ended_at = parse_date_time(ended_at, orders = c("%Y-%m-%d %H:%M:%S", "%Y-%m-%d %H:%M", "%m/%d/%Y %H:%M")))

head(df)
```

### 4.Analysis

Spliting date into day, month and year.

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
trips<-mutate(trips,travel_time=as.double(difftime(ended_at,started_at)/60))

trips$day<- format(as.Date(trips$started_at), "%d")
trips$month<- format(as.Date(trips$started_at), "%m")
trips$year<- format(as.Date(trips$started_at), "%Y")
trips$day_of_week <- format(as.Date(trips$started_at), "%A")


```

#### 4.1 Removing trips when performed quality checks

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
trips_v2 <- trips[!(trips$start_station_name == "HQ QR" | trips$travel_time<0),]

```

#### 4.2 Descriptive analisys
Summarizing data, calculating mean, median, max and min values for the cleaned and filtered dataset.
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
aggregate(trips_v2$travel_time ~ trips_v2$member_casual, FUN = mean)
aggregate(trips_v2$travel_time ~ trips_v2$member_casual, FUN = median)
aggregate(trips_v2$travel_time ~ trips_v2$member_casual, FUN = max)
aggregate(trips_v2$travel_time ~ trips_v2$member_casual, FUN = min)


trips_v2$day_of_week <- ordered(trips_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))

aggregate(trips_v2$travel_time ~ trips_v2$member_casual + trips_v2$day_of_week, FUN = mean)

trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%  #creates weekday field using wday()
  group_by(member_casual, weekday) %>%                  #groups by usertype and weekday
  summarise(number_of_rides = n()							          #calculates the number of rides and average duration 
  ,average_duration = mean(travel_time)) %>% 		        # calculates the average duration
  arrange(member_casual, weekday)			



```

### 5. Vizualizations
Graphs of average trip length for casual users and members per day.
Graphs of numbers of trips per day for casual users and members.

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

# Let's create a visualization for average duration
trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(travel_time)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")+ 
  labs(title = "Averge trip duration",x="Day of the week",y="Minutes")


trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(travel_time)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")+ 
  labs(title = "Total rides per day",x="Day of the week",y="Number of trips")
```


### 6.Export data
Saving summaized data into .csv file
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
counts <- aggregate(trips_v2$travel_time ~ trips_v2$member_casual + trips_v2$day_of_week, FUN = function(x) c(Sum = sum(x), Mean = mean(x)))
write.csv(counts, file = 'avg_ride_length.csv')



```




